# Reproducible Research: Peer Assessment 1
==============================================

## Loading and preprocessing the data

Before we begin, we will use some libraries not included in the base package, this piece of code make sure they are installed correctly.

```{r}
if(require(xtable)) {
  message("Library xtable loaded correctly")
} else {
  message("Library xtable not found, attempting to install...")
  install.packages("xtable")
  if(require(plyr)) {
    message("Library xtable installed and loaded correctly")
  } else {
    stop("Could not install xtable")
  }
}

if(require(plyr)) {
  message("Library plyr loaded correctly")
} else {
  message("Library plyr not found, attempting to install...")
  install.packages("plyr")
  if(require(plyr)) {
    message("Library plyr installed and loaded correctly")
  } else {
    stop("Could not install plyr")
  }
}

if(require(ggplot2)) {
  message("Library plyr loaded correctly")
} else {
  message("Library ggplot2 not found, attempting to install...")
  install.packages("ggplot2")
  if(require(plyr)) {
    message("Library ggplot2 installed and loaded correctly")
  } else {
    stop("Could not install ggplot2")
  }
}
```


For the first part, we will temporarily download the file to the hard drive, and load the dataframe into R and delete the temporary files.

```{r}
## File URL Location

fileurl<- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"

## Download file to a temporary location, read in dataframe, and delete tempfile

temp<-tempfile()
download.file(fileurl, temp)
filename<-unzip(temp, list=TRUE)[[1]]
data<-read.csv(unzip(temp,filename), na.strings = "NA")
unlink(temp)
unlink(filename)
str(data)
```

We can see that the date column is misassigned as a factor, it would be better to classify this as an actual date

```{r}
data$date <- as.Date(data$date, format = "%Y-%m-%d")
str(data)
```

The date seems to be fixed now. Let's look at a summary of the data

```{r,showtable, results="asis"}
xt <- xtable(summary(data))
print(xt, type = "html")
```


## What is mean total number of steps taken per day?

We will use the ddply() function from the plyr package to summarise the data and make a dtotalDF dataframe.

```{r}

dtotalDF <- ddply(
     data, 
     c("date"), 
     summarize, 
     dailytotal = sum(steps, na.rm = TRUE))

head(dtotalDF)
```

It seems to have correctly summed up the dates, so let's make the histogram and calculate mean and median (to the nearest step).

```{r}
mean <- as.integer(mean(dtotalDF$dailytotal))
median <- as.integer(median(dtotalDF$dailytotal))
qplot(data = dtotalDF, x = dailytotal, geom = "histogram", binwidth = 1000)
```

The mean number of steps taken each day is __`r mean`__  
The median number of steps taken each day is __`r median`__   

## What is the average daily activity pattern?

For this part we will use a similar method to summarize the data using the ddply
function.      

This time, we will summarize by interval as the identifier.
     


```{r}
intervalsDF <- ddply(
     data, 
     c("interval"), 
     summarize, 
     mean = mean(steps, na.rm = TRUE))

ggplot(intervalsDF, aes(x = interval, y = mean)) +
     geom_line()

maxmean <- format(
     intervalsDF[intervalsDF$mean==max(intervalsDF$mean, na.rm=TRUE),"mean"],
     digits = 3)

maxinterval <- as.numeric(format(
     intervalsDF[intervalsDF$mean==max(intervalsDF$mean, na.rm=TRUE),"interval"],
     digits = 3))

hours <- as.numeric(trunc(maxinterval/60))
minutes <- as.numeric(format(maxinterval - hours*60, digits = 0))


maxtime <- strptime(paste(hours, minutes, sep = ":"), "%H:%M")
maxtime <- format(maxtime, format = "%I:%M %P")

```

Averaged over all days the 5-minute interval with the most number of steps occurred at __`r maxtime`__, or interval __`r maxinterval`__. <br />
<br />
The number of steps recorded during this time was __`r maxmean`__

## Imputing missing values

First, count the number of rows with NAs in them.


```{r}
head(data)

sum(rowSums(is.na(data)))

```

There are a lot!    

For simplicity sake, we will replace NA values with the mean number of steps for that specific interval accross all days (already computed in the intervalsDF above)  

"NAintervals" is a vector containing what the interval value is for that specific NA value in the steps column.    

"index" is a vector containing the indeces where there is a NA value.

The replacement is carried out by a for loop.


```{r}

imputDF <- data

NAintervals <- imputDF[is.na(imputDF),"interval"]

index <- which(is.na(imputDF))

for (i in 1:length(NAintervals)) {
     imputDF[index[i],"steps"] <- intervalsDF[intervalsDF$interval==NAintervals[i],"mean"]
}

head(imputDF)
```

We will now just copy and paste code from above to redo the histogram above, but using the imputed dataset

```{r}
dtotalDF <- ddply(
     imputDF, 
     c("date"), 
     summarize, 
     dailytotal = sum(steps, na.rm = TRUE))

mean <- as.integer(mean(dtotalDF$dailytotal))
median <- as.integer(median(dtotalDF$dailytotal))
qplot(data = dtotalDF, x = dailytotal, geom = "histogram", binwidth = 1000)

```

Using the imputed dataframe:  
The mean number of steps taken each day is __`r mean`__  
The median number of steps taken each day is __`r median`__   

Since the NAs were biasing the daily totals toward 0 when they were present, when they were replaced with the daily mean for that interval, the daily totals were expected to increase. The net impact of imputing the missing data is that the estimates of total daily number of steps is expected to increase.This is reflected in the increase in the above mean and median.

## Are there differences in activity patterns between weekdays and weekends

For this solution, we need to use the weekdays function. We will assign everything that weekdays() returns as "Saturday" or "Sunday" a weekend, everything else will be a weekday.

Then, we will use ddply() splitting on weekday and interval, to obtain means at each time interval for weekend or weekday.  

Lastly, ggplot will be used, using facet_grid to trellis based on weekend or weekday.


```{r}

imputDF$weektime[weekdays(imputDF$date) %in% c("Saturday", "Sunday")] <- "weekend"
imputDF$weektime[!weekdays(imputDF$date) %in% c("Saturday", "Sunday")] <- "weekday"

weektimeDF <- ddply(
     imputDF, 
     c("weektime", "interval"), 
     summarize, 
     means = mean(steps, na.rm = TRUE))

ggplot(weektimeDF, aes(interval, means)) +
     geom_line() +
     facet_grid(weektime~.) +
     geom_line() +
     xlab("5-Minute Interval") +
     ylab("Mean steps taken")

```
